---
title: "IDAG explanation"
author: Dzjon
date: 09/06/2021
output: html_notebook
---

```{r echo = FALSE, warning=FALSE,message=FALSE}
library(tidyverse)
library(ipcalg)
library(pcalg)
```


# Definition
We will denote a causal effect of $A$ on $Y$ by $\Delta Y_A$.
We refer to a graph including $\Delta Y_A$ as an IDAG.
If there is an interaction between some variable and A, there is a directed
arrow (or path) from this variable to $\Delta Y_A$.
In contrast, effect measure modification only corresponds to an association
between some variable and $\Delta Y_A$. [Nilsson]

## Types of effect modification.
According to [VanderWeele], there are four types of effect modification.
In these examples we have intervention $A$ and outcome variable $Y$.
$Q$ and $A$ have a direct effect on $Y$, and there is an interaction effect
between $Q$ and $A$.

## No effect modification
```{r, echo = F }
set.seed(42)
n <- 1000
alpha <- 0.01
data <- tibble(
  A = rnorm(n),
  Q = rnorm(n),
  Y = A + Q + rnorm(n),
)

pc(suffStat = list(C = cor(data), n = nrow(data)),
  indepTest = gaussCItest,
  alpha = alpha,
  labels = colnames(data)) %>%
  ipc(data, x = 'A', y = 'Y', alpha = alpha) %>%
  plot()
```


### Direct effect modification
```{r, echo = F }
set.seed(42)
n <- 1000
alpha <- 0.01
data <- tibble(
  A = rnorm(n),
  Q = rnorm(n),
  Y = A + Q + A * Q + rnorm(n),
)

pc(suffStat = list(C = cor(data), n = nrow(data)),
  indepTest = gaussCItest,
  alpha = alpha,
  labels = colnames(data)) %>%
  ipc(data, x = 'A', y = 'Y', alpha = alpha) %>%
  plot()
```

We call $Q$ a direct effect modifier for the causal effect of $A$ on $Y$,
because $A$ is a direct cause of $Y$.
This is represented in the IDAG as an arrow from $Q$ to $\Delta Y_A$.

## Indirect effect modification

```{r, echo = F }
set.seed(42)
n <- 1000
alpha <- 0.01
data <- tibble(
  A = rnorm(n),
  C = rnorm(n),
  Q = C + rnorm(n),
  Y = A + Q + A * Q + rnorm(n),
)

pc(suffStat = list(C = cor(data), n = nrow(data)),
  indepTest = gaussCItest,
  alpha = alpha,
  labels = colnames(data)) %>%
  ipc(data, x = 'A', y = 'Y', alpha = alpha) %>%
  plot(2)
```
We say that $C$ is an indirect effect modifier for the causal effect of $A$ on $Y$,
since $C$ affects $Y$ indirectly through $Q$.
This is represented in the IDAG as an arrow from $C$ to $Q$.


## Effect modification by proxy

```{r echo = FALSE}
pc(suffStat = list(C = cor(data), n = nrow(data)),
  indepTest = gaussCItest,
  alpha = alpha,
  labels = colnames(data)) %>%
  ipc(data, x = 'A', y = 'Y', alpha = alpha) %>%
  plot(1)
```

$X$ is an effect modifier for the causal effect of $A$ on $Y$
because $X$ contains information about $Q$ (which serves a direct effect modifier
for the causal effect of $A$ on $Y$).
However, because $R$ is not a cause of $A$, we say that $X$ is an effect
modifier by proxy.
This is represented in the IDAG as an arrow from $X$ to $R$.

## Effect modification by common cause
```{r echo = FALSE}
set.seed(42)
n <- 1000
alpha <- 0.01
data <- tibble(
  A = rnorm(n),
  C = rnorm(n),
  X = C + rnorm(n),
  Q = C + rnorm(n),
  Y = A + Q + A * Q + rnorm(n),
)

pc(suffStat = list(C = cor(data), n = nrow(data)),
  indepTest = gaussCItest,
  alpha = alpha,
  labels = colnames(data)) %>%
  ipc(data, x = 'A', y = 'Y', alpha = alpha) %>%
  plot(2)
```
Because $X$ contains information about $C$ which is a cause of the direct
effect modifier $Q$, we call $X$ an effect modifier by common cause of the
effect of $A$ on $Y$.

## Total effect modification
```{r echo = FALSE}
set.seed(42)
n <- 1000
alpha <- 0.01
data <- tibble(
  A = rnorm(n),
  Q = rnorm(n),
  C = rnorm(n),
  X = A + A * Q + Q + rnorm(n),
  Y = X + C + rnorm(n),
)

pc(suffStat = list(C = cor(data), n = nrow(data)),
  indepTest = gaussCItest,
  alpha = alpha,
  labels = colnames(data)) %>%
  ipc(data, x = 'A', y = 'Y', alpha = alpha) %>%
  plot()
```
```{r echo = FALSE}
set.seed(42)
n <- 1000
alpha <- 0.01
data <- tibble(
  A = rnorm(n),
  B = rnorm(n),
  C = rnorm(n),
  Y = A + B + C + A*B + rnorm(n),
)

pc(suffStat = list(C = cor(data), n = nrow(data)),
  indepTest = gaussCItest,
  alpha = alpha,
  labels = colnames(data)) %>%
  ipc(data, x = 'A', y = 'Y', alpha = alpha) %>%
  plot()
```
