---
title: "IPC algorithm showcase"
author: Dzjon Hessing
date: 08/06/2021
output: html_notebook
---
```{r warning=FALSE,message=FALSE}
library(ipcalg)
library(tidyverse)
```
# Nilsson examples
Here I show that the `ipc` algorithm can generate all the IDAGs in the Nilsson paper from data.
In each case, we are interested in the causal relation between A and Y.
For some examples, the CPDAG contains more than one possible DAG.
In these cases, I have selected the right one by hand.

## Figure 1
An example of a standard directed acyclic graph (DAG) and two possible interaction DAGs (IDAGs)
Variables A (warfarin) and Q (smoking) influence Y (ischaemic stroke).
The first IDAG suggests that Q also influences the effect of A on Y,
whereas the second suggests that this is not the case.

### Interaction effect
```{r}
set.seed(42)
n <- 1000
alpha <- 0.01

data <- tibble(
  A = rnorm(n),
  Q = rnorm(n),
  Y = A + Q + A * Q + rnorm(n),
)

pc(
  suffStat = list(C = cor(data), n = nrow(data)),
  indepTest = gaussCItest,
  alpha = alpha,
  labels = colnames(data)
) %>%
  ipc(data, x = 'A', y = 'Y', alpha = alpha) %>%
  plot()
```

### No interaction effect
```{r}
n <- 1000
data <- tibble(
  A = rnorm(n),
  Q = rnorm(n),
  Y = A + Q + rnorm(n),
)

pc(suffStat = list(C = cor(data), n = nrow(data)),
   indepTest = gaussCItest,
   alpha = alpha,
   labels = colnames(data)) %>%
  ipc(data, x = 'A', y = 'Y',) %>%
  plot()
```

## Figure 2
Confounded interaction or ‘effect modification by proxy’.
A standard directed acyclic graph (DAG) is given in the left panel  and an interaction DAG (IDAG) in the right panel.
Variables X (genotype) and A (bariatric surgery) influence Y (weight loss), with an interaction present.
The effect of A is modified by Q (hair colour), but there is no interaction between A and Q.
```{r}
set.seed(42)
n <- 1000
data <- tibble(
  X = rnorm(n),
  Q = X + rnorm(n),
  A = rnorm(n),
  Y = X + A + A * X + rnorm(n)
)

pc(suffStat = list(C = cor(data), n = nrow(data)),
   indepTest = gaussCItest,
   alpha = alpha,
   labels = colnames(data)) %>%
  ipc(data, x = 'A', y = 'Y', alpha) %>%
  plot(1)
```
## Figure 3
Two examples of standard directed acyclic graphs (DAGs) (left) and two interaction DAGs (IDAGs) (right).
The variable Y (a disease) is directly influenced by A (treatment), Q (smoking) and potentially also X (education).
The first DAG is compatible with the first IDAG ,
whereas the seconds DAG is compatible with either of the IDAGs.
```{r}
set.seed(42)
n <- 1000
data <- tibble(
  X = rnorm(n),
  Q = X + rnorm(n),
  A = rnorm(n),
  Y = Q + A + A * Q + rnorm(n)
)

pc(suffStat = list(C = cor(data), n = nrow(data)),
   indepTest = gaussCItest,
   alpha = alpha,
   labels = colnames(data)) %>%
  ipc(data, x = 'A', y = 'Y', alpha) %>%
  plot(2)
```

## Simple summary
We haven't looked at the `summary` function yet.
Let's create a simple example.
```{r}
set.seed(42)
alpha <- 0.01
n <- 1000
data <- tibble(
  A = rnorm(n),
  B = rnorm(n),
  Y = A - B + 2 * A * B + rnorm(n),
)

ipc.fit <- pc(suffStat = list(C = cor(data), n = nrow(data)),
              indepTest = gaussCItest,
              alpha = alpha,
              labels = colnames(data)) %>%
  ipc(data, x = 'A', y = 'Y', alpha)

plot(ipc.fit)
```
```{r}
summary(ipc.fit)
```

The `summary` function gives an overview of the causal structure.
For each DAG in the CPDAG it shows:

- the estimated formula
- the causal function
- the ACE (as an expression)
- the interaction effects stratified by quartile

## Total effect modification

```{r}
set.seed(42)
n <- 1000
data <- tibble(
  A = rnorm(n),
  B = rnorm(n),
  C = A + B + A * B + rnorm(n),
  D = rnorm(n),
  Y = C + D + rnorm(n)
)

pc(suffStat = list(C = cor(data), n = nrow(data)),
   indepTest = gaussCItest,
   alpha = alpha,
   labels = colnames(data)) %>%
  ipc(data, x = 'A', y = 'Y', alpha) %>%
  plot()
```
The `summary` function shows the derived formulas from the DAG and IDAG, and
the average causal effect as an expression.

```{r}
summary(ipc.fit)
```


## Confounder
This shows an example where the calculated CPDAG gives many possible
combinations.
```{r}
set.seed(42)
n <- 1000
data <- tibble(
  C = rnorm(n),
  A = C + rnorm(n),
  Y = C + A + A * C + rnorm(n)
)

ipc.fit <- pc(
  suffStat = list(C = cor(data), n = nrow(data)),
  indepTest = gaussCItest,
  alpha = alpha,
  labels = colnames(data)) %>%
  ipc(data, x = 'A', y = 'Y', alpha)

plot(ipc.fit$pc.fit, main = "")
plot(ipc.fit)
summary(ipc.fit)
```
## Diamond
```{r}
set.seed(42)
n <- 1000
alpha <- 0.01
data <- tibble(
  A = rnorm(n),
  B = A + rnorm(n),
  D = rnorm(n),
  C = A + D + rnorm(n),
  Y = A + B + C + A * B + rnorm(n)
)

ipc.fit <- pc(
  suffStat = list(C = cor(data), n = nrow(data)),
  indepTest = gaussCItest,
  alpha = alpha,
  labels = colnames(data)) %>%
  ipc(data, x = 'A', y = 'Y', alpha)
plot(ipc.fit)
summary(ipc.fit)
```
